A Reimplementation of Python’s dict using Simple Tabulation Hashing and Linear Probing

### 2012-05-20 docs/paper.tex
More implementation details
More writing

\subsubsection{Optimizations}

After implementation and profiling, we found that dictionary table lookups had
over twice as many level 1 cache misses with tabulation hashing than without.
Hypothesizing that the tabulation tables were evicting other, more relevant
data from the level 1 cache.  To fix this, we inserted x86 memory prefetch
instructions into our hash code to tell the CPU's caching system that it should
deprioritize storing it in cache.  We used GCC's
\begin{tt}\_\_builtin\_prefetch(void \*addr, char rw, char locality)\end{tt}
to do this.  However, this only yielded about a 1\% performance improvement.

@@

For simple tabulation hashing with variable length strings, we decided to use a
fixed number of tables $N$ and round robin between them for every byte.  We
found the optimal value of $N$ to be 8 experimentally.  We limited our choices
to powers of 2 so that costly modulo operations could be avoided by using bit
masks Figure ~\ref{fig:tables} shows the performance of different numbers of
tabulation tables.  Eight turned out to be optimal.
 \begin{figure}[H]
   \centering
   \includegraphics[width=4in]{tables.pdf}
   \caption{6000000 keys}
   \label{fig:tables}
 \end{figure}

Eight tables provided the best performance.

 \begin{figure}[H]
   \centering
   \includegraphics[width=4in]{randstring.pdf}
   \caption{Random string performance}
   \label{fig:randstring}
 \end{figure}
 \begin{figure}[H]
   \centering
   \includegraphics[width=4in]{seqstring.pdf}
   \caption{Sequential string performance}
   \label{fig:seqstring}
 \end{figure}
 \begin{figure}[H]
   \centering
   \includegraphics[width=4in]{chainlength.pdf}
   \caption{Simple tabulation hashing decreases the average chain length}
   \label{fig:seqstring}
 \end{figure}

>>>>

We also experimented with indexing using every pair of bytes in the string and
using $2^16$ tabulation table entries.

To implement linear probing, we modified Python's dictionary code to start at
the table entry defined by the hash value and then search sequentially until it
finds the item it is looking for or an empty cell.
\subsubsection{Optimizations}

After implementation and profiling, we found that dictionary table lookups had
over twice as many level 1 cache misses with tabulation hashing than without.
Hypothesizing that the tabulation tables were evicting other, more relevant
data from the level 1 cache.  To fix this, we inserted x86 memory prefetch
instructions into our hash code to tell the CPU's caching system that it should
deprioritize storing it in cache.  We used GCC's
\begin{tt}\_\_builtin\_prefetch(void \*addr, char rw, char locality)\end{tt}
to do this.  However, this only yielded about a 1\% performance improvement,
negligible compared to the performance improvement of different hash schemes.

@@

For simple tabulation hashing with variable length strings, we decided to use a
fixed number of tables $N$ and round robin between them for every byte.  We
found the optimal value of $N$ to be 8 experimentally.  We limited our choices
to powers of 2 so that costly modulo operations could be avoided by using bit
masks Figure ~\ref{fig:tables} shows the performance of different numbers of
tabulation tables with both 1 and 2 byte indices.  Eight hash tables indexing
every byte separately resulted in the best performance.    We believe this is due
to the low entropy for fewer tables and too much memory bandwidth for a higher
number of tables.  The optimal result may also be an artifact of the fact that
we were using on the order of magnitude of thousands to millions of dictionary
keys.  Our optimal result resulted 256 64
bit table entries in each of 8 tables using 16KB of memory.
 \begin{figure}[H]
   \centering
   \includegraphics[width=4.1in]{tables.pdf}
   \caption{Eight tabulation tables with 1 byte indices provides the best performance for a
   dictionary with 6000000 keys}
   \label{fig:tables}
 \end{figure}

 \begin{figure}[H]
   \centering
   \includegraphics[width=4.1in]{randstring.pdf}
   \caption{Random string performance}
   \label{fig:randstring}
 \end{figure}
 \begin{figure}[H]
   \centering
   \includegraphics[width=4.1in]{seqstring.pdf}
   \caption{Sequential string performance}
   \label{fig:seqstring}
 \end{figure}
 \begin{figure}[H]
   \centering
   \includegraphics[width=4.1in]{chainlength.pdf}
   \caption{Simple tabulation hashing decreases the average chain length}
   \label{fig:seqstring}
 \end{figure}
